A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-04-18 response.md

Response for Submission 66, ICFP 2019
======

We thank the reviewers for their helpful comments and positive reception of the paper.
We answer main questions below.

Review A
------
<!--
One minor point of criticism I have is that there appear to be some unexplained differences between the presentation in the paper and the Abella mechanization, which made it harder than necessary for me to relate the two and to follow some of the proof sketches in the paper. For example, the (binary) multi-step reduction relations on algorithmic and declarative worklists seem to be represented by unary predicates (judge and dc respectively) in the Abella script, which look more like big-step than small-step reduction relations. This change of representation may even impact the structure of some proofs (e.g. the induction hypothesis used in the proofs of soundness and completeness). I know very little about Abella, so this choice of representation may well be more idiomatic than the one presented in the paper. Still, I would have liked to see a short discussion of the pros/cons of the two representations and why different representations are used in the paper and the proof script. More extensive documentation of the proof script (e.g. more comments) would also be helpful in navigating the mechanization.
-->

> The difference between the presentation in the paper and the Abella mechanization.

In the paper, the main algorithmic rules are written in a simple small step style
with no premise on the reduction relation (-->) itself.
And the rules are not overlapping with each other.
These properties indicates that a trivial translation to
the big step style used in the proofs preserves equivalence.

We believe the presentation of the algorithmic rules in the paper is much clear
and restricted to the form $\Gamma$ --> $\Gamma$,
while the big step unary relation in the proof simplfies the induction scheme.
More importantly, the difference is just a matter of presentation,
but the logic behind remains unchanged.

> The proof scripts are hard to read, and some parts are inconsistent with the paper.

We apologize for the lack of clean-up on the proof scripts.
We are now refactoring the scripts,
after that we will provide a better documentation on the encoding details
that might cover all the concerns by the reviewer,
preferrably to the appendix of the paper.

Review B
------
<!--
On the negative part, I am not sure that one can, from the paper, infer general hints/guidelines about the formalization of the correctness/completeness/... of an inference algorithm for a complex typing discipline. This is mainly due to the fact that the authors concentrate their attention to one single inference algorithm.
-->
<!--
Tom:
- Apparently Reviewer C believes they will be able to try out our approach on new problems.
- I also think there is an obvious generalization to other bidirectional type systems.
- We can say that we will extend the discussion to point this out.
-->

> General hints/guidlines about the formalization of the properties of an inference algorithm for a complex typing descipline.

We believe that the algorithm proposed can adapt to other bidirectional type systems,
and we will further extend the discussion to point this out.

Review C
------
<!--
In the course of their investigation, the authors found several problems with the mathematical proof DK give and provided alternatives. I am not in a position to comment on whether these criticisms of DK are justified, nor do I find this significant for the evaluation of the paper.
-->
<!--
Jimmy:
If the reviewer are not interested in the criticisms,
we can reduce relevant discussion in the paper.
My point was that I am not sure whether DK's algorithm and proofs
could be easily formalized, but it seems no easy.
And I decided not to take time investigate,
given that there are already some issues, though they might be minor.
-->

> Criticisms of DK's proof

...

### 2019-04-18 response.md

Response for Submission 66, ICFP 2019
======

We are grateful for the helpful reviews and are elated that you appreciate our work.
In what follows  we respond to your main comments.

### 2019-04-18 response.md

We apologize for the lack of clean-up on the proof scripts.
We are now refactoring the scripts,
after that we will provide a better documentation on the encoding details
that might cover all the concerns by the reviewer,
preferrably to the appendix of the paper.

Review B
------
<!--
On the negative part, I am not sure that one can, from the paper, infer general hints/guidelines about the formalization of the correctness/completeness/... of an inference algorithm for a complex typing discipline. This is mainly due to the fact that the authors concentrate their attention to one single inference algorithm.
-->
<!--
Tom:
- Apparently Reviewer C believes they will be able to try out our approach on new problems.
- I also think there is an obvious generalization to other bidirectional type systems.
- We can say that we will extend the discussion to point this out.
-->

> General hints/guidlines about the formalization of the properties of an inference algorithm for a complex typing descipline.

That is a good point. We believe that our approach can indeed be adapted to other bidirectional type systems, and we will extend the discussion in the paper to point this out.
We are also happy to see that Reviewer C already looks forward to trying out our approach in other settings.

### 2019-04-18 response.md

Review A
------
<!--
One minor point of criticism I have is that there appear to be some unexplained differences between the presentation in the paper and the Abella mechanization, which made it harder than necessary for me to relate the two and to follow some of the proof sketches in the paper. For example, the (binary) multi-step reduction relations on algorithmic and declarative worklists seem to be represented by unary predicates (judge and dc respectively) in the Abella script, which look more like big-step than small-step reduction relations. This change of representation may even impact the structure of some proofs (e.g. the induction hypothesis used in the proofs of soundness and completeness). I know very little about Abella, so this choice of representation may well be more idiomatic than the one presented in the paper. Still, I would have liked to see a short discussion of the pros/cons of the two representations and why different representations are used in the paper and the proof script. More extensive documentation of the proof script (e.g. more comments) would also be helpful in navigating the mechanization.
-->

> The difference between the presentation in the paper and the Abella mechanization.

In the paper, the main algorithmic rules are written in a simple small step style
with no premise on the reduction relation (-->) itself.
And the rules are not overlapping with each other.
These properties indicates that a trivial translation to
the big step style used in the proofs preserves equivalence.

We believe the presentation of the algorithmic rules in the paper is much clearer
and restricted to the form $\Gamma$ --> $\Gamma$,
while the big step unary relation in the proof simplfies the induction scheme.
More importantly, the difference is just a matter of presentation,
but the logic behind remains unchanged.

@@

> The proof scripts are hard to read, and some parts are inconsistent with the paper.

We apologize for the state of the proof scripts. Currently, we are refactoring them.
and will document the mapping between the paper and the mechanisation.

### 2019-04-18 response.md

Response for Submission 66, ICFP 2019
======

We are grateful for the helpful reviews and are happy
that you appreciate our work. In what follows we
respond to your main comments.

@@

Review A
------
<!--
One minor point of criticism I have is that there appear to be some
unexplained differences between the presentation in the paper and the
Abella mechanization, which made it harder than necessary for me to
relate the two and to follow some of the proof sketches in the
paper. For example, the (binary) multi-step reduction relations on
algorithmic and declarative worklists seem to be represented by unary
predicates (judge and dc respectively) in the Abella script, which
look more like big-step than small-step reduction relations. This
change of representation may even impact the structure of some proofs
(e.g. the induction hypothesis used in the proofs of soundness and
completeness). I know very little about Abella, so this choice of
representation may well be more idiomatic than the one presented in
the paper. Still, I would have liked to see a short discussion of the
pros/cons of the two representations and why different representations
are used in the paper and the proof script. More extensive
documentation of the proof script (e.g. more comments) would also be
helpful in navigating the mechanization.
-->

> The difference between the presentation in the paper and the Abella mechanization.

In the paper, the main algorithmic rules are written in a simple small
step style with no premise on the reduction relation (-->) itself.
And the rules are not overlapping with each other.  These properties
indicates that a trivial translation to the big step style used in the
proofs preserves equivalence.

We believe the presentation of the algorithmic rules in the paper is
much clearer and restricted to the form $\Gamma$ --> $\Gamma$, while
the big step unary relation in the proof simplifies the induction
scheme.  More importantly, the difference is just a matter of
presentation, but the logic behind remains unchanged.
We will mention this difference in presentation clearly in the paper.

> The restriction of subtyping to simple  types is effectively an
> equivalence (and thus symmetric).  Is this correct?

Yes, this is correct. Note also that DK's algorithm also suggests a
similar symmetric unification.

@@

> The proof scripts are hard to read, and some parts are inconsistent with the paper.

We apologize for the state of the proof scripts. Currently, we are refactoring them.
and will document the mapping between the paper and the mechanisation
in the final version of the paper.

@@

Review B
------
<!--
On the negative part, I am not sure that one can, from the paper, infer general hints/guidelines about the formalization of the correctness/completeness/... of an inference algorithm for a complex typing discipline. This is mainly due to the fact that the authors concentrate their attention to one single inference algorithm.
-->
<!--
Tom:
- Apparently Reviewer C believes they will be able to try out our approach on new problems.
- I also think there is an obvious generalization to other bidirectional type systems.
- We can say that we will extend the discussion to point this out.
-->

> General hints/guidlines about the formalization of the properties of
> an inference algorithm for a complex typing descipline.

### 2019-04-18 response.md

> The restriction of subtyping to simple  types is effectively an
> equivalence (and thus symmetric).  Is this correct?

Yes, this is correct. Note also that DK's algorithm also suggests a
similar symmetric unification.

> The proof scripts are hard to read, and some parts are inconsistent with the paper.

We apologize for the state of the proof scripts. Currently, we are refactoring them,
and will document the mapping between the paper and the mechanisation
in the final version of the paper.

> How does polynomial complexity follow from the reduction rule being deterministic?

We did not justify the point clearly in the paper.
A detailed proof will be provided for the final version.

> Why not treat the languages of declarative and algorithmic worklists as subsets?

The proof script allows one to treat them as subsets.
However, we explicitly treat them as different syntactic sorts in the paper
to avoid potential confusion.

> I could not find a direct counterpart of this theorem in the Abella script.

The `decidable_thm` Theorem is the one that proves the Termination Theorem (4.12),
following what is described from L918,
and the `decidable` Theorem is its pretty version.
We will make the theorem names consistent for publish.

### 2019-04-19 response.md

Review A
------

> The difference between the presentation in the paper and the Abella mechanization.

In the paper, the main algorithmic rules are written in a simple small
step style with no premise on the reduction relation (-->) itself.
And the rules are not overlapping with each other.  These properties
indicates that a trivial translation to the big step style used in the
proofs preserves equivalence.

We believe the presentation of the algorithmic rules in the paper is
clearer and restricted to the form $\Gamma$ --> $\Gamma$, while
the big step unary relation in the proof simplifies the induction
scheme.  More importantly, the difference is just a matter of
presentation, but the logic behind remains unchanged.
We will mention this difference in presentation clearly in the paper.

### 2019-04-19 response.md

> The restriction of subtyping to simple  types is effectively an
> equivalence (and thus symmetric).  Is this correct?

Yes, this is correct. Note also that DK's algorithm also suggests a
similar symmetric unification.

> The proof scripts are hard to read, and some parts are inconsistent with the paper.

We apologize for the state of the proof scripts. Currently, we are refactoring them,
and will document the mapping between the paper and the mechanisation
in the final version of the paper.

> How does polynomial complexity follow from the reduction rule being deterministic?

We did not justify the point clearly in the paper.
A detailed proof will be provided for the final version.

> Why not treat the languages of declarative and algorithmic worklists as subsets?

The proof script allows one to treat them as subsets.
However, we explicitly treat them as different syntactic sorts in the paper
to avoid potential confusion.

> I could not find a direct counterpart of this theorem in the Abella script.

The `decidable_thm` Theorem is the one that proves the Termination Theorem (4.12),
following what is described from L918, and the `decidable` Theorem is its pretty version.
We will make the theorem names consistent in the final version.

### 2019-04-20 response.md

Review A
------

> The difference between the presentation in the paper and the Abella mechanization.

In the paper, the main algorithmic rules are written in a simple small
step style with no premise on the reduction relation (-->) itself.
And the rules are not overlapping with each other.  These properties
indicates that a trivial translation to the big step style used in the
proofs preserves equivalence.

We believe the presentation of the algorithmic rules in the paper is
clearer and restricted to the form $\Gamma$ --> $\Gamma'$, while
the big step unary relation in the proof simplifies the induction
scheme.  More importantly, the difference is just a matter of
presentation, but the logic behind remains unchanged.
We will mention this difference in presentation clearly in the paper.

@@

> The proof scripts are hard to read, and some parts are inconsistent with the paper.

We apologize for the state of the proof scripts. Currently, we are refactoring them,
and will document the mapping between the paper and the mechanisation
in the final version of the paper.

> How does polynomial complexity follow from the reduction rule being deterministic?

We did not justify the point clearly in the paper.
We will either provide a detailed proof in the final version, or omit
this informal claim.

@@

> I could not find a direct counterpart of this theorem in the Abella script.

The `decidable_thm` Theorem is the one that proves the Termination Theorem (4.12),
following what is described from L918, and the `decidable` Theorem is its pretty version.
We will make the theorem names consistent in the final version.

Review B
------

> General hints/guidelines about the formalization of the properties of
> an inference algorithm for a complex typing discipline.

### 2019-05-17 main.tex
Not Abstract, sorry
macros in main.tex that helps generate parts
Modification according to comments and known typos

@@

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been \emph{manually} proven
\emph{sound}, \emph{complete} and \emph{decidable}. We present 
a mechanical formalization in the Abella theorem prover 
of DK's declarative type system with a novel algorithmic system. We have
a few reasons to use a new algorithm. Firstly, our new algorithm
employs \emph{worklist judgments}, which allow precise capture of the
scope of variables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing and some
incorrect proofs,
which complicate the task of writing a formal proof. 
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. Since such type inference algorithms are quite
subtle and have a complex metatheory, mechanical
formalizations are an important advance in type-inference research.
\end{abstract}

>>>>

% Appendix only
%\usepackage[28-]{pagesel}
% Main article with references
%\usepackage[-27]{pagesel}

@@

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been \emph{manually} proven
\emph{sound}, \emph{complete} and \emph{decidable}. We present 
a mechanical formalization in the Abella theorem prover 
of DK's declarative type system with a novel algorithmic system. We have
a few reasons to use a new algorithm. Firstly, our new algorithm
employs \emph{worklist judgments}, which allow precise capture of the
scope of variables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing and some
incorrect proofs,
which complicate the task of writing a machanized proof. 
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. Since such type inference algorithms are quite
subtle and have a complex metatheory, mechanical
formalizations are an important advance in type-inference research.
\end{abstract}

### 2019-05-17 sections/Metatheory.tex
newpage before appendix; move measures to appendix; new ==>> symbol
Modification according to comments and known typos

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
Given \emph{wf }$\Gm$ containing no judgment chains:
\begin{enumerate}
    \item If $\Om \vdash A' \le B'$ and $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$
        \\then $\Gm \Vdash A \le B \redto \nil$;
    \item If $\Om \vdash e \Lto A'$ and $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$
        \\then $\Gm \Vdash e \Lto A \redto \nil$;
    \item If $\Om \vdash e \To A$ and $\Gm \Vdash e \To_a 1 \le 1 \sto \Om \Vdash e \To_a 1 \le 1$
        \\then $\Gm \Vdash e \To_a 1 \le 1 \redto \nil$;
    \item If $\Om \vdash \appInf{A'}{e}{C}$ and
        $\Gm \Vdash \appInfAlg{A}{e}[a][1 \to 1] \sto \Om \Vdash \appInfAlg{A'}{e}[a][1 \to 1]$
        \\then $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \redto \nil$.
\end{enumerate}
\end{corollary}

@@

Finally, we show that our algorithm terminates. In other words:
\begin{theorem}[Termination]
Given a well-formed input $\Gm$, the algorithm terminates.
\end{theorem}
Our termination proof is based on a lexicographic group of induction measures\\
$\langle |\Gm|_e, |\Gm|_\Leftrightarrow, |\Gm|_\forall, |\Gm|_{\al}, |\Gm|_\to + |\Gm| \rangle$
on the worklist $\Gm$. The definitions of these measures can be found in Figure~\ref{fig:measures}. The first two, 
$|\cdot|_e$ and $|\cdot|_\Leftrightarrow$, measure the total size of terms
and the total difficulty of judgments, respectively. In the latter, check judgments
count for 2, inference judgments for 1 and function inference judgments for 3.
Other two measures, $|\cdot|_\forall$ and $|\cdot|_\to$, count the total number of
universal quantifiers and function types, respectively. Finally,
$|\cdot|_{\al}$ counts the number of existential variables in the worklist,
and $|\cdot|$ is simply the length of the worklist.

\begin{figure}[t]
\hfill \framebox{$|\Gm|_e$} \hfill  \llap{Size measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_e &= \textstyle\sum_{e\Lto A\in\Gm}|e|_e + \sum_{e\To_a \jg\in\Gm}|e|_e +
    \sum_{\appInfAlg{A}{e}\in\Gm}|e|_e\\
\text{where } |x|_e &= |()|_e = 1 \qquad |\lam e|_e = |e|_e + 1\\
|e_1~e_2|_e &= |e_1|_e + |e_2|_e + 1 \qquad |e:A|_e = |e| + 1
\end{aligned}
\end{gather*}
\hfill \framebox{$|\Gm|_\Leftrightarrow$} \hfill  \llap{Judgment measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_\Leftrightarrow &= 2\cdot\#_{e\Lto A\in\Gm} +
    \#_{e\To_a \jg\in\Gm} + 3\cdot\#_{\appInfAlg{A}{e}\in\Gm}
\end{aligned}
\end{gather*}
\hfill \framebox{$|\Gm|_\forall$} \hfill  \llap{Polymorphism measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_\forall &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\forall + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\forall +
    \sum_{A\le B\in\Gm} |A|_\forall + |B|_\forall\\
\text{where } |1|_\forall &= |a|_\forall = |\al|_\forall = 1\\
|A\to B|_\forall &= |A|_\forall + |B|_\forall \qquad |\all A|_\forall = |A|_\forall + 1
\end{aligned}
\end{gather*}
\hfill \framebox{$|\Gm|_\to$} \hfill  \llap{Function measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_\to &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\to + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\to +
    \sum_{A\le B\in\Gm} |A|_\to + |B|_\to\\
\text{where } |1|_\to &= |a|_\to = |\al|_\to = 1\\
|A\to B|_\to &= |A|_\to + |B|_\to + 1 \qquad |\all A|_\to = |A|_\to
\end{aligned}
\end{gather*}
\hfill \framebox{$|\Gm|_{\al}$} \hfill  \llap{Existential measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_{\al} &= \#_{\al\in\Gm}\\
\end{aligned}
\end{gather*}
\caption{Worklist measures}
\label{fig:measures}
\end{figure}

@@

>>>>

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
Given \emph{wf }$\Gm$ containing no judgment chains:
\begin{enumerate}
    \item If $\Om \vdash A' \le B'$ and $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$
        \\then $\Gm \Vdash A \le B \redto \nil$;
    \item If $\Om \vdash e \Lto A'$ and $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$
        \\then $\Gm \Vdash e \Lto A \redto \nil$;
    \item If $\Om \vdash e \To A$ and $\Gm \Vdash e \To_a 1 \le 1 \sto \Om \Vdash e \To_a 1 \le 1$
        \\then $\Gm \Vdash e \To_a 1 \le 1 \redto \nil$;
    \item If $\Om \vdash \appInf{A'}{e}{C}$ and
        $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \sto \Om \Vdash \appInfAlg{A'}{e}[a][1 \le 1]$
        \\then $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \redto \nil$.
\end{enumerate}
\end{corollary}

@@

Finally, we show that our algorithm terminates:
\begin{theorem}[Termination]
Given a well-formed input $\Gm$, the algorithm terminates.
\end{theorem}
Our termination proof is based on a lexicographic group of induction measures\\
$\langle |\Gm|_e, |\Gm|_\Leftrightarrow, |\Gm|_\forall, |\Gm|_{\al}, |\Gm|_\to + |\Gm| \rangle$
on the worklist $\Gm$. Formal definitions of these measures can be found in
Figure~\ref{fig:measures} in the Appendix. The first two, 
$|\cdot|_e$ and $|\cdot|_\Leftrightarrow$, measure the total size of terms
and the total difficulty of judgments, respectively. In the latter, check judgments
count for 2, inference judgments for 1 and function inference judgments for 3.
Another two measures, $|\cdot|_\forall$ and $|\cdot|_\to$, count the total number of
universal quantifiers and function types, respectively. Finally,
$|\cdot|_{\al}$ counts the number of existential variables in the worklist,
and $|\cdot|$ is simply the length of the worklist.

@@

\jimmy{Discuss briefly on the difference of encodings in the script and this paper.}
While the algorithmic rules are in a small-step style,
the proof script rewrites them into big-step style,
where inductions are more strightforward to handle.
In addition, we do prove the equivalence of the two representations.
